---
title: "R Coding Workshop: 3rd Meeting"
subtitle: "GIS & Geospatial Data Analysis (Fall 2025)"
execute:
  echo: true
format:
  html:
    output-file: "w5-idx.html"
    css: styles.css
    toc: true
    df-print: kable
  revealjs:
    output-file: "w5-slides.html"
    css: styles.css
    df-print: kable
    slide-number: true
    scrollable: true
    code:
      echo: true
      eval: true
---

## R Coding Workshop: 3rd Meeting

## Outline for today

- Install and load R Packages
- *Tidyverse*: coding style and collection of packages
  - Pipeline operator
  - `readr`: Read and write data
  - Tidy Data: prepared for data analysis (next week)
- Exploratory Data Analysis:`dplyr` continued
- `ggplot`: Data Visualization

## Tidyverse

Tidyverse coding style and packages
```{r}
library(tidyverse)
```

## Read and write data

```{r}
uni_df <- tibble::tribble(
  ~university, ~year, ~current_country, ~lat, ~lon, ~exist_today,
  "Paris", 1150, "France", 48.8566, 2.3522, TRUE,
  "Salerno", 1173,  "Italy", 40.7711, 14.7905, TRUE,
  "Reggio", 1188, "Italy", 44.6450, 10.9277, TRUE,
  "Oxford", 1190, "United Kingdom", 51.7520, -1.2576, TRUE,
  "Bologna", 1200, "Italy", 44.4989, 11.3275, TRUE
)
uni_df
```

```{r}
uni_fpath <- 'data/university-1200.csv'
uni_df |> write_csv(uni_fpath)
```

```{r}
uni_df |> rm()
```

```{r}
uni_df <- read_csv(uni_fpath)
uni_df |> class()
```

```{r}
library(sf)
library(mapview)
```
```{r}
uni_sf <- uni_df |> st_as_sf(coords = c('lon','lat'), crs = 4326)
uni_sf |> mapview(label = 'university')
```

```{r}
uni_sf_fpath <- 'data/uni-1200-sf.gpkg'
# uni_sf |> st_write(uni_sf_fpath)
```

```{r}
uni_sf |> rm()
uni_sf <- st_read(uni_sf_fpath)
uni_sf |> mapview(label = 'university')
```

## `dplyr` for Exploratory Data Analysis

- More functions from `dplyr`
- Merits of pipeline operator `|>`

## Tidyverse EDA

**Import Data: Challenged Books from 2000 to 2010**

- source: American Library Society (America Library Association)
```{r}
library(bayesrules)

book_challenge_df <- book_banning |> as_tibble()
book_challenge_df |> head()
```

::: {.callout-note appearance="minimal" collapse="true" title="Tidy Data and Data Cleaning (next week)"}

```{r}
book_challenge_df |> summary()
```
```{r}
book_challenge_df |> dim()
```

- remove 11 books with missing values (require justification!)

```{r}
book_challenge_df <- book_challenge_df |> drop_na()
book_challenge_df |> dim()
```

- duplicated rows?
```{r}
# book_challenge_df |> group_by(book_id) |> filter(n_distinct(title) > 1) |> arrange(book_id) |> tail()
```

```{r}
# book_challenge_df |> distinct(book_id, title)
```

:::

Dimension:

- (`row` x `column`)
- 920 Observations: number of recorded book challenges
- 17 Variables:
  - observation-level
    - *title*, *author*, *date*
  - state-level variables
    - *median_income*, *political_value_index*

Summary statistics:

- the *data type* and the *value range* of the dataset
```{r}
book_challenge_df |> summary()
```

## `dplyr` functions: `mutate()`

- `mutate()`: creating new variables (columns) from existing variables
- case: typecasting

```{r}
book_challenge_df <- book_challenge_df |>
  mutate(
    yr = year(date),
    month = month(date),
    day = day(date),
    week_day = wday(date, label = TRUE)
    )
```

```{r}
all(book_challenge_df$year == as.numeric(book_challenge_df$yr))
```

```{r}
# book_challenge_df <- book_challenge_df |> mutate(year = yr) |> select(-yr)
```
- `count()`
```{r}
book_challenge_df |> count(week_day)
```

## `dplyr` functions: `group_by()`

- group observations by like variable value, why?

```{r}
nrow(book_challenge_df[book_challenge_df$removed == 1,]) / nrow(book_challenge_df)
```

## `dplyr` functions: `group_by()`

- subset, stratify
```{r}
book_challenge_df |> group_by(year) |> summarize(
    total_count = n(),
    removed_count = sum(removed == 1),
    removed_pct = (removed_count / total_count) * 100
)
```

```{.r}
# A tibble: 11 × 4
    year total_count removed_count removed_pct
   <dbl>       <int>         <int>       <dbl>
 1  2000          71            26        36.6
 2  2001         111            26        23.4
 3  2002          52            16        30.8
 4  2003          88            14        15.9
 5  2004          83            20        24.1
 6  2005          52            15        28.8
 7  2006          54             9        16.7
 8  2007          38            15        39.5
 9  2008         122            37        30.3
10  2009         197            22        11.2
11  2010          52            14        26.9
```

- expressive pipeline that aligns with DS workflow!
- Base R:
```{.r}
total_count <- as.vector(
    tapply(
        book_challenge_df$removed,
        book_challenge_df$year,
        length
        )
)
removed_count <- as.vector(
    tapply(
        book_challenge_df$removed == 1,
        book_challenge_df$year,
        sum
        )
)
year_summary <- data.frame(
    year = 2000:2010,
    total_count = total_count,
    removed_count = removed_count,
    removed_pct = (removed_count / total_count) * 100
)
year_summary
```

## Group Data by Geounit

```{r}
state_level_vars <- c(
    "state", "political_value_index",
    "median_income", "hs_grad_rate", "college_grad_rate"
    )
# book_challenge_df |> group_by(across(state_level_vars)) |> summarize(count = n())
```

```{r}
state_summary <- book_challenge_df |>
  group_by_at(state_level_vars) |>
  summarize(count = n())

state_summary
```

```{r}
# sanity check
state_summary |> dim() |> print()
state_summary$count |> sum()
```

## Adding Spatial Dimension

```{r}
library(sf)
library(mapview)
library(rnaturalearth)
library(viridis)
```

- Prepare the Geometry
  
```{r}
state_sf <- ne_states(country = 'United States of America') |> select(postal, gn_name, gadm_level, region)
state_sf |> dim() |> print()
state_sf |> mapview()
```

- Combine by column value match

```{r}
plot_sf <- state_sf |>
  left_join(
    state_summary,
    by = join_by(postal == state)
)
plot_sf |> head()
```

```{r}
state_summary |> select(state, count) |> filter(state == 'WA')
```

## Choropleth

```{r}
plot_sf |> mapview(zcol = 'count', label = 'gn_name')
```

```{r}
plot_sf |> filter(is.na(count))
```

## `ggplot` plot with more controls

```{r}
plot_sf |> ggplot() + geom_sf(aes(fill = count), color = 'white', size = 0.01) + scale_fill_viridis() + theme_bw()
```